---
title: "Analysis examples in curatedPCaData"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Analysis examples in curatedPCaData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r packages}
library(curatedPCaData)
library(survival)
library(survminer)
```

# Example analyses with curatedPCaData

## Brief overview to MultiAssayExperiment functions

## Survival analysis

### Kaplan-Meiers

#### Gleason-score

#### Genomic risk

### Cox proportional hazard models

## Risk scores

### Decipher vs. BCR

### Oncotype DX vs. BCR

### Prolaris vs. BCR

## Purity scores

### Pathology purity vs. DeMixT

## Multi-omics

### Copy number deletion and mutation combined

### Multivariable regression from multiple assay types



# Session info

```{r session}
sessionInfo()
```

