---
title: "Example correlations for genes in curatedPCaData"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Example correlations for genes in curatedPCaData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
```

```{r packages}
library(curatedPCaData)
library(MultiAssayExperiment)
library(RaggedExperiment)
library(survival)
library(survminer)
library(corrplot)
```

### General trends in gene expression across datasets

Using the curatedpcadata package, patterns in gene expression across different datasets can be studied.

```{r correlation_plot, eval = FALSE}

calculate_correlation<-function(dataset,gene){
  name=deparse(substitute(dataset))
  data_subset=as.data.frame(t(dataset[gene,]))
  
  data_subset$sample=rownames(data_subset)
  data_subset[,1]=as.numeric(data_subset[,1])
  
  dataset=as.data.frame(t(dataset))
  
  
  cor_values=numeric(0)
  
  for( i in 1:ncol(dataset)){
    indvidual_cor_values=print(cor(dataset[,i],data_subset[,1],method = "spearman"))
    cor_values <- c(cor_values, indvidual_cor_values) 
  }
  
  cor_matrix=data.frame(matrix(NA, nrow = ncol(dataset), ncol = 2))
  colnames=colnames(dataset)[1:ncol(dataset)]
  colnames=as.vector(colnames)
  cor_matrix[,1]=colnames
  cor_matrix[,2]=cor_values
  #paste0("spearman_correlation_coeff_",deparse(substitute(dataset)))
  colnames(cor_matrix)=c("gene",name)
  return(cor_matrix)
}


correlation_of_correlations<-function(gene){
  tcga=as.data.frame(curatedPCaData::mae_tcga[["gex.fpkm"]])
  taylor=as.data.frame(curatedPCaData::mae_taylor[["gex.rma"]])
  abida=as.data.frame(curatedPCaData::mae_abida[["gex.relz"]])
  ren=as.data.frame(curatedPCaData::mae_ren[["gex.relz"]])
  barbieri=as.data.frame(curatedPCaData::mae_barbieri[["gex.relz"]])
  igc=as.data.frame(curatedPCaData::mae_igc[["gex.rma"]])
  friedrich=as.data.frame(curatedPCaData::mae_friedrich[["gex.logq"]])
  barwick=as.data.frame(curatedPCaData::mae_barwick[["gex.logq"]])
  chandran=as.data.frame(curatedPCaData::mae_chandran[["gex.rma"]])
  icgcca=as.data.frame(curatedPCaData::mae_icgcca[["gex.rma"]])
  kim=as.data.frame(curatedPCaData::mae_kim[["gex.rma"]])
  kunderfranco=as.data.frame(curatedPCaData::mae_kunderfranco[["gex.logr"]])
  sun=as.data.frame(curatedPCaData::mae_sun[["gex.rma"]])
  true=as.data.frame(curatedPCaData::mae_true[["gex.logr"]])
  wallace=as.data.frame(curatedPCaData::mae_wallace[["gex.rma"]])
  wang=as.data.frame(curatedPCaData::mae_wang[["gex.rma"]])
  weiner=as.data.frame(curatedPCaData::mae_weiner[["gex.rma"]])
  
  cor_matrix_tcga = calculate_correlation(tcga,gene)
  cor_matrix_taylor= calculate_correlation(taylor,gene)
  cor_matrix_abida= calculate_correlation(abida,gene)
  cor_matrix_ren= calculate_correlation(ren,gene)
  cor_matrix_barbieri= calculate_correlation(barbieri,gene)
  cor_matrix_igc= calculate_correlation(igc,gene)
  cor_matrix_friedrich= calculate_correlation(friedrich,gene)
  cor_matrix_barwick= calculate_correlation(barwick,gene)
  cor_matrix_chandran= calculate_correlation(chandran,gene)
  cor_matrix_icgcca= calculate_correlation(icgcca,gene)
  cor_matrix_kim= calculate_correlation(kim,gene)
  cor_matrix_kunderfranco= calculate_correlation(kunderfranco,gene)
  cor_matrix_sun= calculate_correlation(sun,gene)
  cor_matrix_true= calculate_correlation(true,gene)
  cor_matrix_wallace= calculate_correlation(wallace,gene)
  cor_matrix_wang= calculate_correlation(wang,gene)
  cor_matrix_weiner= calculate_correlation(weiner,gene)
  
  
  
  #put all data frames into list
  df_list <- list(cor_matrix_tcga, cor_matrix_taylor, cor_matrix_abida,cor_matrix_ren,
                  cor_matrix_barbieri,cor_matrix_igc,cor_matrix_friedrich,cor_matrix_barwick,
                  cor_matrix_chandran,cor_matrix_icgcca,cor_matrix_kim,
                  cor_matrix_kunderfranco,cor_matrix_sun,cor_matrix_true,cor_matrix_wallace,
                  cor_matrix_wang,cor_matrix_weiner)
  
  combined_cor_list=Reduce(function(x, y) merge(x, y, all=F), df_list)  
  rownames(combined_cor_list)=combined_cor_list[,"gene"]
  combined_cor_list=combined_cor_list[,-1]
  
  combined_cor_list[,c(1:17)]=sapply( combined_cor_list[,c(1:17)], as.numeric )
  return(combined_cor_list)
}
```

For example the correlation of AR expression to the expression of all the other genes across all datasets can be studied with the help of a correlation plot as shown:
  
```{r, results=FALSE, eval = FALSE}

library(corrplot)
combined_cor_list=correlation_of_correlations("AR")
corr_of_corr = cor(combined_cor_list)
```

In order to save time for vignette illustration purposes, above correlation matrix has been precomputed and stored as a `.RData` workspace.

```{r load_workspace}
load("correlation_between_gex.RData")
```

The corresponding correlation plot:

```{r, dpi=150, fig.width=7, fig.height=7, out.width="100%"}
corrplot::corrplot(corr_of_corr,method = 'color')

```

# Session info

```{r session}
sessionInfo()
```
