---
title: "getting-started"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)
library(tidyverse)
library(viridis)
```


This vignette introduces `curatedPCaData` and provides an overview of the  package using data from TCGA as an example. 

***

3 different data types are currently available: 

1. `gex_*.RData` contains gene expression values
2. `cna_*.RData` contains copy number values
3. `clinical_*.RData` contains clinical/phenotype information
    
These elements are created with internal functions within `curatedPCaData` and are then used to derive additional measures, such as immune scores. 
Not all datasets will contain each of these elements but at  minimum will contain `clinical_*.rda` and `gex_*.rda`.

These individual elements are combined into a `MultiAssayExperiment object` (abbreviation: `MAE`, in lower case by convention) called `mae_*.rda`. 


## Introduction 

The `curatedPCaData` package contains a collection of manually curated datasets concerning patients diagnosed with prostate cancer. The datasets within this package have followed uniform processing and naming conventions to allow users to more easily reproduce similar analyses between datasets and spend less time concerned with harmonzing data from different resources. 

In this vignette, we will : 

1. Provide an introduction to `curatedPCaData`. 
2. Show how to run a simple analysis using the data 

## Importing data

The individual elements are housed within the `data-raw/` folder of the package [github repo](https://github.com/Syksy/curatedPCaData) and follows a simple naming convention of `data type` and `first author` separated by a `_` (see dataset types above). 
Each indivdual element for creating the final `MAE` were created using internal functions (accessible with `:::`).
A `MAE` for each study is provided to the users. 

To get a full list of available datasets use the `data` function:
```{r importing}
library(curatedPCaData)
data(package="curatedPCaData")
```

Individual datasets can also be imported by using the `data` function

```{r data-import}
data("mae_tcga")
mae_tcga
```
For more information on working with `mae`s please see the BioConductor `MutliAssayExperiment` package [documentation](https://bioconductor.org/packages/devel/bioc/vignettes/MultiAssayExperiment/inst/doc/QuickStartMultiAssay.html). 

## Example analysis

Daniel will put some text here about doing a simple analysis and perform an analysis here: 
  + survival analysis usinng alternative outcomes (BCR or mets)
  + use multiple datasets
  + example: http://www.bioconductor.org/packages/release/data/experiment/vignettes/curatedOvarianData/inst/doc/curatedOvarianData_vignette.pdf

```{r subset_data,eval=FALSE}
wide_tcga <- MultiAssayExperiment::wideFormat(mae_tcga["PTEN", , ],
    colDataCols=c("overall_survival_status", "days_to_overall_survival", "gleason_grade"))
summary(glm(days_to_overall_survival ~ GEX_PTEN + gleason_grade, data = wide_tcga))
```

## Immune deconvolution

Tumor progression depends on the immune cell composition in the tumor microenvironment. The '[immunedeconv](https://github.com/icbi-lab/immunedeconv)' package consists of different computational methods to computationally estimate immune cell content using gene expression data.

In this package, we provide estimates of immune cell content from the following deconvolution methods:

1. quantiseq
2. xcell
3. epic
4. mcp_counter
5. cibersortx

The estimates from each of these methods are stored in the MAE object as a seperate assay as shown for example in the Taylor dataset
```{r}
mae_taylor
```

To access the quantiseq results for the Taylor et. al dataset, the following command can be used
```{r}
mae_taylor[["quantiseq"]]
```

Similarly to access results from the other methods, the following can be run
```{r}
mae_taylor[["xcell"]]
mae_taylor[["epic"]]
mae_taylor[["mcp_counter"]]
mae_taylor[["cibersortx"]]

```

Each row of the deconvolution matrix represents the content of a certain immune cell type and the columns represent the patient sample IDs.























