---
title: "Clinical metadata sweep across datasets"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{Clinical metadata sweep across datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  cache = TRUE
)

library(curatedPCaData)

template <- curatedPCaData::template_prad

# Sweep across the different rows available in PRAD template
sweeps <- list()

# List of MAE objects
maes <- grep("mae_", utils::data(package="curatedPCaData")$result[,"Item"], value=TRUE)

# Create functions for summarizing MAE-object fields
summarize_feature <- function(mae, feature){
	if(feature %in% colnames(MultiAssayExperiment::colData(mae))){
		x <- MultiAssayExperiment::colData(mae)[,feature]
	}else{
		return(NA)
	}
	# Summarizing character-fields
	if(all(is.na(x))){
		"N/A"
	}
	else if(template[which(template$col.name == feature),"var.class"] %in% c("character", "string")){
		# Tabulate and order tabulation
		tab <- table(x, useNA="ifany")
		tab <- sort(tab, decreasing=TRUE)
		# If there are only 3 or less unique character string entries summarize them
		if(length(unique(x))<=3){
			return(paste0(paste0(paste0(names(tab), ": ", tab), " (", round(tab/sum(tab)*100,0), "%)"), collapse=";"))
		}
		# Else summarize top 3 common entries and NAs
		else{
			return(paste0(paste0(paste0(paste0(names(tab)[1:3], ": ", tab[1:3]), " (", round(tab[1:3]/sum(tab)*100,0), "%)"), collapse=";"), "; ..."))
		}		
	# Summarizing numeric fields
	}else if(template[which(template$col.name == feature),"var.class"] %in% c("integer", "numeric")){
		# Cast to numeric; for example, an error entry may be psa value ">2000" causing string value
		x <- as.numeric(x)
		# If there are only 3 or less unique numeric entries summarize them tabulated
		if(length(unique(x))<=3){
			# Tabulate and order tabulation
			tab <- table(x, useNA="ifany")
			tab <- sort(tab, decreasing=TRUE)
			return(paste0(paste0(paste0(names(tab), ": ", tab), " (", round(tab/sum(tab)*100,0), "%)"), collapse=";"))
		}
		# For 4 or more unique values, use quantiles
		else{
			# Contains missing values
			if(any(is.na(x))){
				return(paste0("[",paste0(round(quantile(x, na.rm=TRUE),2), collapse=","),"], NA n=", sum(is.na(x))))
			# No missing values
			}else{ 
				return(paste0("[",paste0(round(quantile(x),2), collapse=", "),"]"))
			}
		}		
	}
	else{
		return("Unidentified var.class")
	}
}
missingness_feature <- function(mae, feature){
	x <- MultiAssayExperiment::colData(mae)[,feature]
	if(sum(is.na(x))>0){
		paste0("n=", sum(is.na(x)), " (", round(100*sum(is.na(x))/length(x),0), "%)")
	}
	else{
		"-"
	}
}

sweep <- function(var){ 

	m <- matrix(nrow=length(maes), ncol=3)
	colnames(m) <- c("Values", "Class", "NAs (%)")
	rownames(m) <- maes

	for(maename in maes){
		eval(parse(text=paste0("mae <- curatedPCaData::", maename)))

		m[maename,"Values"] <- summarize_feature(mae, feature = var)

	}
	m
}		


# Create result list
for(varname in template[,"col.name"]){
	print(varname)
	sweeps[[length(sweeps) + 1]] <- sweep(var = varname)
}
names(sweeps) <- template[,"col.name"]

```

# Metadata for curatedPCaData studies

curatedPCaData version ```packageVersion("curatedPCaData")``` contain ```length(grep("MAE", utils::data(package="curatedPCaData")$result))``` datasets, with a total of ```nrow(curatedPCaData::template_prad)``` clinical template candidate variables as outlined in the metadata template `curatedPCaData::template_prad`. Note that this contains multiple missing values, and `NA` values indicate that the values were either not available or could not be reliable inferred.

# Iteration over variables in PRAD template

```{r}
sweeps
```

# Session info

```{r session}
sessionInfo()
```


