---
title: "TCGA-example"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

This vignette introduces `curatedPCaData` and provides an overview of the  package using data fro TCGA as an example. 

***

3 different types of datasets are currently available: 
    + `_gex.rda` contains gene expression values
    + `_cna.rda` contains copy number values
    + `_clinical.rda` contains clinical/phenotype information
    
The TCGA data included in the package contains all 3 of these. 

## Introduction 

In this vignette, we will : 

1. Examine the data as provided in `curatedPCaData`
2. Format the data  
3. Calculate ... 
4. Run a simple analysis 

## The data 

All datasets are currently housed within the `data/` folder of the package and follows a simply naming convention of `first-author` and `data type` separated by a `_` (see dataset types above).

For more information on the various datasets and the values within them please see the [data dictionaries vignette](vignettes/data-dictionaries.Rmd). 

```{r importing}
library(curatedPCaData)
dim(tcga_clinical)
```

## Formatting 

Individual data pieces can be transformed into an MAE for easier handeling using the `create_mae()` function. The function only takes one argument - the prefix for the dataset. 

```{r transform}

tcga_mae <- create_mae(study_name = "TCGA")

```

