---
title: "TCGA-example"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCGA-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)

```

This vignette introduces `curatedPCaData` and provides an overview of the  package using data from TCGA as an example. 

***

3 different data types are currently available: 
    + `gex_*.RData` contains gene expression values
    + `cna_*.RData` contains copy number values
    + `clincial_*.RData` contains clinical/phenotype information
    
These elements are created with internal functions within `curatedPCaData` and are then used to derive additional measures, such as immune scores. 

These individual elements are combined into a `MultiAssayExperiment object` (abbreviation: `MAE`, in lower case by convention) called `mae_*.rda`. 


## Introduction 

In this vignette, we will : 

1. Examine the data as provided in `curatedPCaData`. 
2. Run a simple analysis 

## The data 

The individual elements are housed within the `data-raw/` folder of the package and follows a simply naming convention of `data type` and `first author` separated by a `_` (see dataset types above). 
A `MAE` for each study is provided to the users under the `data/` folder. 

```{r importing}
library(curatedPCaData)
mae_tcga
```

All functions and indivdual elements for creating the final `MAE` were created using internal functions (accessible with `:::`) and saved in the `data-raw/` folder in the github and is not compiled with the package.

## Simple analysis

```{r subset_data}

wide_tcga <- MultiAssayExperiment::wideFormat(mae_tcga["PTEN", , ],
    colDataCols=c("overall_survival_status", "days_to_overall_survival", "gleason_grade"))

summary(glm(days_to_overall_survival ~ GEX_PTEN + gleason_grade, data = wide_tcga))


```



